---
title: "Cyclistic Case Study"
author: "Kent Ward"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(skimr)
library(readxl)
library(dplyr)
library(scales)
library(patchwork)
library(viridis)
```

# 1. Load the Required Libraries

These packages will help with data import, manipulation, and visualization.

```{r}
library(tidyverse)
library(lubridate)
library(skimr)
library(readxl)
library(dplyr)
library(scales)
library(patchwork)
library(viridis)
```



# 2. Import Data

We import the 2019 and 2020 Q1 bike trip datasets directly from the local `raw_data` folder.

```{r}
all_trips_2019 <- read_excel("raw_data/Raw_Divvy_Trips_2019_Q1.xlsx")
all_trips_2020 <- read_excel("raw_data/Raw_Divvy_Trips_2020_Q1.xlsx")
```

# 3. Inspect the Structure of Both Datasets

Before cleaning, we review data types and contents.

```{r}
datasets <- list("2019" = all_trips_2019, "2020" = all_trips_2020)
lapply(datasets, str)
lapply(datasets, skim)
lapply(datasets, summary)
```

# 4. Clean and Standardize Column Formats

We trim whitespace and convert all character data to lowercase for consistency.

```{r}
datasets <- lapply(datasets, function(df) {
  df %>% mutate(across(where(is.character), ~ trimws(tolower(.))))
})
```

# 5. Rename Columns for 2019 Dataset

This ensures both datasets have the same structure before merging.

```{r}
datasets[["2019"]] <- datasets[["2019"]] %>%
  rename(
    ride_id = trip_id,
    rideable_type = bikeid,
    started_at = start_time,
    ended_at = end_time,
    start_station_name = from_station_name,
    start_station_id = from_station_id,
    end_station_name = to_station_name,
    end_station_id = to_station_id,
    member_casual = usertype
  )
```

# 6. Standardize Data Types

Convert ride_id and rideable_type to characters in 2019 so they match 2020.

```{r}
datasets[["2019"]] <- datasets[["2019"]] %>%
  mutate(
    ride_id = as.character(ride_id),
    rideable_type = as.character(rideable_type)
  )
```

# 7. Combine the Two Datasets

```{r}
all_trips <- bind_rows(datasets[["2019"]], datasets[["2020"]])
```

# 8. Drop Unnecessary Columns

Remove irrelevant or inconsistent fields across years.

```{r}
all_trips <- all_trips %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, tripduration))
```

# 9. Recode Membership Types

Standardize `subscriber` as `member` and `customer` as `casual`.

```{r}
all_trips <- all_trips %>%
  mutate(member_casual = recode(member_casual, "subscriber" = "member", "customer" = "casual"))
```

# 10. Add Time Columns

Break the datetime into components for better grouping later.

```{r}
all_trips <- all_trips %>%
  mutate(
    date = as.Date(started_at),
    month = format(date, "%m"),
    day = format(date, "%d"),
    year = format(date, "%Y"),
    day_of_week = wday(started_at, label = TRUE, abbr = FALSE)
  )
```

# 11. Calculate Ride Length in Seconds

```{r}
all_trips$ride_length <- as.numeric(difftime(all_trips$ended_at, all_trips$started_at, units = "secs"))
```

# 12. Remove Bad or Invalid Data

Filter out rides with negative time or system test locations like "HQ QR".

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "hq qr" | all_trips$ride_length < 0),]
```

# 13. Descriptive Summary

```{r}
summary(select(all_trips_v2, ride_length))
```

# 14. Compare Ride Length by Member Type

Use both aggregate and `dplyr` summary for clean outputs.

```{r}
ride_summary_overall <- all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(
    `Average Ride (min)` = round(mean(ride_length, na.rm = TRUE), 0),
    `Median Ride (min)` = round(median(ride_length, na.rm = TRUE), 0),
    `Max Ride (min)` = round(max(ride_length, na.rm = TRUE), 0),
    `Min Ride (min)` = round(min(ride_length, na.rm = TRUE), 0)
  )
ride_summary_overall
```

# 15. Weekly Usage and Ride Behavior

```{r}
# Create new column for weekday, then summarize ride count and average duration
weekly_usage <- all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE, abbr = FALSE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(
    number_of_rides = n(),
    average_duration = round(mean(ride_length, na.rm = TRUE), 0),
    .groups = "drop"
  ) %>%
  arrange(member_casual, weekday)
```

# 16. Visualize: Rides and Duration by Day

```{r}
# Plot: Average Number of Rides
plot_rides <- ggplot(weekly_usage, aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(option = "plasma", begin = 0.2, end = 0.9) +
  labs(
    title = "Ride Count by Day",
    x = "Day of the Week",
    y = "Number of Rides",
    fill = "Rider Type"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Plot: Average Ride Duration
plot_duration <- ggplot(weekly_usage, aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(option = "plasma", begin = 0.2, end = 0.9) +
  labs(
    title = "Avg. Ride Duration",
    x = "Day of the Week",
    y = "Duration (seconds)",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#Combined plots 
```{r}
combined_plot <- (plot_rides + plot_duration) +
  plot_layout(widths = c(1, 1), guides = "collect") +
  plot_annotation(title = "Ride Count & Duration by Rider Type") &
  theme(legend.position = "bottom")

combined_plot
```

# 17. Save Plot for Presentation

```{r}
ggsave("ride_usage_combined.png", plot = combined_plot, width = 10, height = 6, dpi = 300)

ggsave("ride_count_by_day.png", plot = plot_rides, width = 8, height = 5, dpi = 300)

ggsave("avg_ride_duration.png", plot = plot_duration, width = 8, height = 5, dpi = 300)
```
